---
title: "Data visualization hands-on session"
author: "Alan Lorenzetti"
date: "8/11/2020"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(DT)
```

# Introduction

In this session we will try to cover a few topics on data visualization.

1. We are going to reproduce a protein-protein interaction network I've made for a paper published recently. For that, we are going to use Cytoscape.
2. We will explore different types of variables and plot them using the R language environment. More specifically we will use RStudio Cloud and a package called `tidyverse`.

# Cytoscape

## Installing Cytoscape and add-ins (or plugins)

## Data structure

## Plotting a protein-protein interaction network

### Context

### Execution

# Visualization in R using the `tidyverse` package

## Installing `tidyverse` package

```{r install}
#install.packages("tidyverse")
library(tidyverse)

haloExp = read_tsv("data/haloExpression.tsv")
```

## Variable types

### Discrete variables

* Counts:  

  + The number of times you cursed this damn virus this week (5)
  + Number of stars in the Milky Way (250 billion or 2.5 x 10^11)
  + Number of genes in a genome (2600)
  + Number of messenger RNA molecules in a bacterial cell (~1800) [Moran et al. 2013]
  + Length of a messenger RNA (1573 nucleotides)

### Continuous variables

* Measures:

  + Height of an individual (162.56 centimeters) 
  + Weight of an individual (89.77 kilograms)

* Ratios:  

  + GC content, i.e., frequency of Guanine or Cytosine bases in a transcript (e.g. 973/1573 = 0.619)

### Categorical variables

* Characteristics:  
  + Color of eyes (black, brown, blue, green, red)
  + Transcript type (protein coding, transporter RNA, ribosomal RNA)

## Data structure

```{r dataStructure, echo=F}
datatable(haloExp)
```

## Visualizations

### Using a single variable

#### Histograms

```{r histogram}
ggplot(data = haloExp, mapping = aes(x = length)) +
  geom_histogram()
```

#### Density curves

```{r density}
ggplot(data = haloExp, mapping = aes(x = GC)) +
  geom_density()
```

#### Bar plots

```{r bar}
ggplot(data = haloExp, mapping = aes(x = biological_class)) +
  geom_bar() +
  coord_flip()
```

#### Boxplots



### Using two variables

#### Scatter plots

```{r scatter}
ggplot(data = haloExp, mapping = aes(x = mRNA_expression, y = protein_expression)) +
  geom_point()
```


### Using three variables

### Using multiple variables

#### 3D scatter plots

#### Heat maps

### Exercises

# Resources

* [RStudio IDE cheatsheet](https://github.com/rstudio/cheatsheets/raw/master/rstudio-ide.pdf)
* [`ggplot2` cheatsheet](https://github.com/rstudio/cheatsheets/raw/master/data-visualization-2.1.pdf) 

# References 

MORAN, M. A. et al. Sizing up metatranscriptomics. The ISME Journal, v. 7, n. 2, p. 237–243, fev. 2013.